---
output: 
  github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Boltzmann–Lotka–Volterra Interaction Model

<!-- badges: start -->

[![R-CMD-check](https://github.com/fabrice-rossi/blvim/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/fabrice-rossi/blvim/actions/workflows/R-CMD-check.yaml) [![Codecov test coverage](https://codecov.io/gh/fabrice-rossi/blvim/graph/badge.svg)](https://app.codecov.io/gh/fabrice-rossi/blvim)

<!-- badges: end -->

`blvim` implements A. Wilson's **Boltzmann–Lotka–Volterra (BLV) interaction model**. 
The model is described in 
[Wilson, A. (2008), "Boltzmann, Lotka and Volterra and spatial structural evolution: an integrated methodology for some dynamical systems", J. R. Soc. Interface, 5:865–871](http://dx.doi.org/10.1098/rsif.2007.1288).

The package's primary goal is to provide a fast implementation of the BLV 
model, complete with a collection of tools designed to explore the results 
through statistical summaries and graphical representations. The secondary goal 
is to facilitate the systematic assessment of how model parameters impact 
the results, again using summaries and graphical representations 
(see `vignette("grid")` for details on this aspect).

## Installation

You can install the development version of `blvim` from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("fabrice-rossi/blvim")
```

## Spatial interaction models

**Spatial interaction models** aim to estimate flows between locations, such as 
workers commuting from residential zones to employment zones. The `blvim` 
package focuses on the maximum entropy models developed by Alan Wilson. 
See `vignette("theory")` for the theoretical background.

In practice, if we have $n$ origin locations and $p$ destination locations, the
goal is to compute a flow matrix $(Y_{ij})_{1\leq i\leq n, 1\leq j\leq p}$, where
$Y_{ij}$ is the flow from origin $i$ to destination $j$. This computation relies 
on characteristics of the origin and destination locations, along with a matrix 
of exchange difficulties, known as a **cost matrix**, 
$(c_{ij})_{1\leq i\leq n, 1\leq j\leq p}$. For example, $c_{ij}$ can represent 
the distance between origin $i$ and destination $j$.

## Usage

The package is loaded in a standard way.
```{r}
library(blvim)
```

### Input data

To compute a spatial interaction model with `blvim`, you need at least a cost 
matrix. In this example, we use the distance between some random locations.

```{r}
set.seed(42)
## random origin locations
origins <- matrix(runif(2 * 30), ncol = 2)
## random destination locations
destinations <- matrix(runif(2 * 10), ncol = 2)
## cost matrix
full_costs <- as.matrix(dist(rbind(origins, destinations)))
cost_matrix <- full_costs[1:nrow(origins), (nrow(origins) + 1):(nrow(origins) + nrow(destinations))]
```

Additionally, since we focus on **production-constrained models**, we must 
specify the production for each origin location (a vector of positive values 
$(X_i)_{1\leq i\leq n}$). Here, we assume a common unitary production.

```{r}
X <- rep(1, nrow(origins))
```

Finally, the simple **static** model requires an attractiveness value for each 
destination location, a vector of positive values $(Z_j)_{1\leq j\leq p}$. We 
again assume a common unitary attractiveness.

```{r}
Z <- rep(1, nrow(destinations))
```

### Static models
In Wilson's production-constrained maximum entropy model, the flows are given
by

$$
Y_{ij} = \frac{X_iZ_j^{\alpha}\exp(-\beta c_{ij})}{\sum_{k=1}^pZ_k^{\alpha}\exp(-\beta c_{ik})},
$$

where $\alpha$ is a return-to-scale parameter and $\beta$ is the inverse of a
cost scale parameter. Note that the flow matrix is **production-constrained**, 
meaning that the total outgoing flow from any origin location is equals 
the production of that location:

$$
\forall i,\quad X_i=\sum_{j=1}^{p}Y_{ij}.
$$

The model is obtained using the `static_blvim()` function:
```{r}
a_model <- static_blvim(cost_matrix, X, alpha = 1.1, beta = 2, Z)
a_model
```

Several functions are provided to extract parts of the result. In particular
`flows()` returns the flow matrix $Y$. 
```{r}
a_model_flows <- flows(a_model)
```
which can be displayed using, for instance, the `image()` function.
```{r a_flow}
#| fig.height: 9
#| fig.width: 3
#| out.width: 25%
#| fig.align: center
par(mar = rep(0.1, 4))
image(t(a_model_flows),
  col = gray.colors(20, start = 1, end = 0),
  axes = FALSE,
  frame = TRUE
)
```

In this representation, each row shows the flows from one origin location to all
destination locations. The package also provides a `ggplot2::autoplot()` 
function, which can be used as follows:

```{r a_flow_ggplot2}
#| fig.height: 9
#| fig.width: 4
#| out.width: 40%
#| fig.align: center
library(ggplot2)
autoplot(a_model, "full") +
  scale_fill_gradient(low = "white", high = "black") +
  coord_fixed()
```



```{r}
b_model <- static_blvim(cost_matrix, X, alpha = 1.1, beta = 15, Z)
b_model
```

```{r b_flow}
#| fig.height: 9
#| fig.width: 4
#| out.width: 40%
#| fig.align: center
autoplot(b_model) +
  scale_fill_gradient(low = "white", high = "black") +
  coord_fixed()
```

As the two figures above exemplify, different values of the parameters $\alpha$ 
and $\beta$ result in more or less concentrated flows.


### Dynamic models

A. Wilson's **Boltzmann–Lotka–Volterra (BLV) interaction model** builds upon the
production-constrained maximum entropy model. The core idea is to update the 
attractiveness of the destination locations based on their incoming flows.

Ideally, we aim for the following condition to hold in the limit:

$$
Z_j =\sum_{i=1}^{n}Y_{ij}, 
$$

where the flows are given by the equations above. The model is estimated using
the `blvim()` function as follows.

```{r}
a_blv_model <- blvim(cost_matrix, X, alpha = 1.1, beta = 2, Z)
a_blv_model
```

Notice that we start with some initial values of the attractiveness, but the 
final values are different. These final values can be obtained using the 
`attractiveness()` function (and visualised here using a bar plot).


```{r a_blv_Z}
#| fig.height: 4
#| fig.width: 8
#| out.width: 80%
#| fig.align: center
par(mar = c(0.1, 4, 1, 0))
a_final_Z <- attractiveness(a_blv_model)
barplot(a_final_Z)
```
In this particular example, one destination location acts as a global attractor 
of all the flows. This pattern is also visible in the final flow matrix.

```{r a_blv_flow}
#| fig.height: 9
#| fig.width: 4
#| out.width: 40%
#| fig.align: center
autoplot(a_blv_model) +
  scale_fill_gradient(low = "white", high = "black")
```

The `autoplot()` function can also be used to show the destination flows or the
attractivenesses values:
```{r a_blv_Z_ggplot2}
#| fig.height: 4
#| fig.width: 8
autoplot(a_blv_model, "attractiveness")
```

Naturally, the results are strongly influenced by the parameters, as shown in 
this second example.

```{r}
b_blv_model <- blvim(cost_matrix, X, alpha = 1.1, beta = 15, Z)
b_blv_model
```

```{r b_blv_Z}
#| fig.height: 4
#| fig.width: 8
autoplot(b_blv_model, "attractiveness")
```

```{r b_blv_flow}
#| fig.height: 9
#| fig.width: 4
#| out.width: 40%
#| fig.align: center
autoplot(b_blv_model) +
  scale_fill_gradient(low = "white", high = "black")
```
