---
title: "Theoretical Background"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Theoretical Background}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(blvim)
```

## Static Models

### Spatial Interaction Models

Spatial interaction models aim to estimate flows between locations, for instance, 
workers commuting from residential zones to employment zones. Models are built from:

-   A collection of $n$ origin locations described by some characteristics 
$(X_i)_{1\leq i\leq n}$
-   A collection of $p$ destination locations described by some characteristics 
$(Z_j)_{1\leq j\leq p}$
-   A collection of $n\times p$ characteristics of the difficulty of travelling 
(in a broad sense) from origin $i$ to destination $j$, $c_{ij}$

The goal is to estimate flows $(Y_{ij})_{1\leq i\leq n, 1\leq j\leq p}$ from 
origin locations to destination locations. A typical hypothesis is for the flows 
to depend on characteristics as follows:

$$
Y_{ij}=f(X_i, Z_j, c_{ij}),
$$

for a well-chosen function $f$. The most well-known spatial interaction model is 
the so-called *gravity model*, which takes the form:

$$
Y_{ij}\propto \frac{X_iZ_j}{c_{ij}^2},
$$

where $\propto$ means proportional to, $c_{ij}$ is supposed to be the distance 
between origin $i$ and destination $j$, and the characteristics $X_i$ and $Z_j$ 
are assumed to be numerical.

### Constraints

Spatial interaction models can be instantiated with different constraints:

-   On production: the $(X_i)$ are interpreted as production values. Then the
model is *production-constrained* if $\sum_{j=1}^pY_{ij}=X_i$ for all $i$;
-   On capacity: the $(Z_j)$ are interpreted as destination capacities. Then the 
model is *capacity-constrained* if $\sum_{i=1}^pY_{ij}=Z_j$ for all $j$;
-   On both: the *doubly-constrained* models are both production- and 
capacity-constrained. One of the most common models used in practice is the 
doubly-constrained gravity model.

### Maximum Entropy Models

In the late 1960s, Alan Wilson developed a collection of spatial interaction 
models based on a maximum entropy principle. In those models, the flow is given by:

$$
Y_{ij} \propto X_iZ_j^{\alpha}\exp(-\beta c_{ij}),
$$

where $\alpha$ and $\beta$ are two parameters interpreted as follows:

-   $\alpha$ is a *return to scale* parameter: if $Z_j$ grows above $1$, its 
actual attractiveness can increase in a super-linear way ($\alpha>1$) or in a 
sub-linear way ($\alpha<1$);
-   $\beta$ acts as the *inverse of the scale of the costs* $c_{ij}$. If those 
costs are distances, for instance, $\frac{1}{\beta}$ can be seen as a cut-off distance.

As with other spatial interaction models, maximum entropy models are constrained. 
In `blvim`, we focus on production-constrained models and thus we have:

$$
Y_{ij} = \frac{X_iZ_j^{\alpha}\exp(-\beta c_{ij})}{\sum_{k=1}^pZ_k^{\alpha}\exp(-\beta c_{ik})},
$$

which ensures:

$$
\forall i,\ \sum_{j=1}^pY_{ij}=X_i.
$$

This model is implemented in `blvim` by the `static_blvim()` function.

## Dynamic Models

The models described above are *static*: they correspond to a frozen view of the 
exchanges with no retroaction of the flow received at one destination location 
on its attractiveness, for instance.

In the 1960s and 1970s, Harris and Wilson developed dynamic models that assume a form of 
etroaction, initially in terms of an equilibrium state.

### Harris and Wilson Equilibrium Values

The main idea from Harris (1964) is that if we interpret incoming flows as 
revenues, then the capacity/attractiveness of a destination location should be 
adapted to those revenues (for production-constrained models). Essentially, this 
could be written as:

$$
\forall j,\ \kappa_j Z_j= D_j,
$$

where the $\kappa_j$ are destination specific conversion factors between capacities 
and costs, and where $D_j$ is the total flow incoming in $Z_j$, i.e.:

$$
\forall j,\  D_j=\sum_{i=1}^nY_{ij}.
$$

Applied to the maximum entropy production-constrained model, this leads to a 
spatial interaction model defined implicitly as follows. We assume given the 
production constraints $(X_i)_{1\leq i\leq n}$ and the cost matrix 
$(c_{ij})_{1\leq i\leq n, 1\leq j\leq p}$, then the attractiveness $(Z_j)_{1\leq j\leq p}$ 
is a solution of the following collection of fixed-point equations:

$$
\forall j,\  \kappa_j Z_j=\sum_{i=1}^n\left(\frac{X_iZ_j^{\alpha}\exp(-\beta c_{ij})}{\sum_{k=1}^pZ_k^{\alpha}\exp(-\beta c_{ik})}\right),
$$

where $\alpha$, $\beta$, and $(\kappa_j)_{1\leq j\leq p}$ are fixed parameters.

From the $(Z_j)_{1\leq j\leq p}$, the flows are computed using the standard 
production-constrained maximum entropy model.

### A Dynamic Model: The Boltzmann–Lotka–Volterra Model

The collection of fixed-point equations can be solved using different techniques, 
for instance, Newton's method. The simplest solution is probably to perform fixed-point 
iterations, as discussed in Harris & Wilson (1978). We start with some initial 
values of the $Z$s, denoted $(Z^{0}_j)_{1\leq j\leq p}$, and then iterate:

$$
\forall j,\ Z_j^{t+1} := \frac{1}{\kappa_j}\sum_{i=1}^n\left(\frac{X_i{(Z^{t}_j)}^{\alpha}\exp(-\beta c_{ij})}{\sum_{k=1}^p{(Z^{t}_k)}^{\alpha}\exp(-\beta c_{ik})}\right),
$$

until convergence.

This iterative scheme can also be interpreted as a dynamic model in which the 
attractiveness of a location is updated to match its incoming flow. The update 
is not instantaneous, and thus an update parameter is introduced, $\epsilon$. 
Then we start again with some initial values of the $Z$s and then iterate:

$$
\begin{align*}
\forall j,\ D_j^{t} &:= \sum_{i=1}^n\left(\frac{X_i{(Z^{t}_j)}^{\alpha}\exp(-\beta c_{ij})}{\sum_{k=1}^p{(Z^{t}_k)}^{\alpha}\exp(-\beta c_{ik})}\right),\\
\forall j,\ Z_j^{t+1} &:= Z_j^t +\epsilon\left(\frac{1}{\kappa_j} D^{t}_j-Z^{t}_j\right).
\end{align*}
$$

Using $\epsilon<1$, typically a small value of $0.01$, gives a full trajectory 
for the $Z$s with the added value of preventing oscillation behaviour.

A variant of the dynamic model has been proposed by Wilson in 2008, replacing 
the second update by a quadratic version:

$$
\forall j,\ Z_j^{t+1} := Z_j^t +\epsilon\left(\frac{1}{\kappa_j} D^{t}_j-Z^{t}_j\right)Z^{t}_j.
$$

Both solutions are implemented in `blvim()`.

### Diversity

A typical behaviour of the equilibrium solutions is a concentration of the
attractiveness on a subset of the destination locations. This can be measured by
the `diversity()` function. It uses the notion of diversity outlined in, for
example, Jost (2006). The main idea of this notion, rephrased in the context of
locations, is to find from the $Z$s an equivalent number of locations, say $m$,
for which the attractiveness would be evenly spread (that is, proportional to $\frac{1}{m}$).

If we normalise the $(Z_j)_{1\leq j\leq p}$ such that they sum to 1, we get the
probability distribution $\mathcal{P}=(p_j)_{1\leq j\leq p}$ over $\{1, \ldots, p\}$.
One way to interpret the notion of diversity is to measure how "random"
$\mathcal{P}$ is. Indeed, if it is concentrated on 1, for instance, with $p_1\simeq 1$,
then samples generated in an i.i.d. manner from $\mathcal{P}$ will not be diverse,
as they will mostly consist of the value 1. Then an *entropy* is a rather natural
diversity index. For instance, the Shannon entropy $H(\mathcal{P})=-\sum_{j}p_j\log p_j$
provides a good diversity index. Based on the principle outlined above, the
diversity of $\mathcal{P}$ is the number $m$ such that the diversity index of
$\mathcal{P}$ is equal to the diversity index of the uniform distribution on
$\{1, \ldots, m\}$. It can be shown that for the Shannon entropy, the diversity of
$\mathcal{P}$ is $\exp(H(\mathcal{P}))$. This holds also for all the Rényi entropies.

Notice that in `diversity()` the diversities are computed according to the incoming
flows, i.e., not using the $(Z_j)_{1\leq j\leq p}$ but rather the $D_j$ defined
above by
$$
\forall j,\  D_j=\sum_{i=1}^nY_{ij}.
$$


## Self-Exchange Configurations

In the general case, exchanges in spatial interaction models follow a bipartite
scheme: production locations send their productions to destination locations.
These locations are two distinct sets. The non-bipartite case is when production
and destination locations are identical. This does not change anything in the
definition of the models (static and dynamic), but new tools can be defined
specifically for this particular case.

### Location Domination and Terminals

These tools are based on notions proposed by J. D. Nystuen and M. F. Dacey (1961).
As any location is both a production and a destination location, one can compare
the outgoing flows to the incoming ones. In particular, Nystuen and Dacey focus
on the maximum output flow and on the total incoming flow (the $(D_j)_{1\leq j\leq p}$).
More precisely, they define the importance of location $j$ as $D_j$ and introduce
the notion of *domination*: a location $k$ is *dominated* if its importance is smaller
than the importance of the location to which it is sending its largest flow. In
equation:

$$
D_k < D_{\arg\max_{j} Y_{kj}}.
$$

A *terminal* is a location that is not dominated. Rihll and Wilson (1991) propose a
relaxed version in which a terminal is a location that receives a total incoming
flow larger than all its individual outgoing flows. Thus, location $k$ is dominated
according to this definition if:

$$
D_k < \max_{j}Y_{kj}.
$$

It is easy to see that a terminal in the Nystuen and Dacey definition is also a
terminal for Rihll and Wilson, but the reverse is false. The functions
`terminals()` and `is_terminal()` compute the terminals according to either
definition. In addition, the number of terminals can be seen as a diversity measure
and is thus included in the definitions supported by `diversity()`.

### Domination Graphs

In addition to extracting terminals from a non-bipartite model, Nystuen and Dacey
propose to analyse the so-called *nodal structure* of the flows. It consists of
the *nodal flows*, i.e. the links between each non-terminal location and the
location to which it is sending its largest flow. This defines a graph on the
locations which is very sparse (it has $p-t$ edges, where $p$ is the number of
locations and $t$ the number of terminals). The function `nd_graph()` computes
this graph according to both terminal definitions.

## Potential function from Osawa et al (2025)

Before going further we introduce simplified notation for the HW model :

$$
w_{ij}=w_{ij}(\beta)=\exp(-\beta c_{ij})
\qquad
S_i=S_i(\mathbf{Z},\alpha,\beta)=\sum_{k=1}^pZ_k^{\alpha}w_{ik}
\qquad
P_{ij}=\frac{Z_j^{\alpha}w_{ij}}{S_i}
$$
so that $P_{ij}$ is the share of origin $i$'s demand attracted to destination $j$.
The total flow received by destination $j$ is

$$
D_j(\mathbf{Z},\alpha,\beta) = \sum_{i=1}^n X_i P_{ij},
$$
and the Harris--Wilson fixed-point residual is

$$
G_j(\mathbf{Z},\alpha,\beta) = D_j(\mathbf{Z},\alpha,\beta) - \kappa_j Z_j,
$$
where $\kappa_j > 0$ is the capacity cost at destination $j$.
A **Harris--Wilson equilibrium** is a vector $\mathbf{Z}^* \geq 0$ satisfying
$G(\mathbf{Z}^*) = \mathbf{0}$.

### The potential function

In 2025 the team of researcher composed by Osawa, Akamatsu and Kogure introduced
a new concept: the HW model can be seen as a large population potential game.
This concept, which comes from game theory, allows us to see the HW model as a
single maximisation problem characterised by a potential function defined as:

$$
f(\mathbf{Z},\alpha,\beta)=A(\mathbf{Z},\alpha,\beta)-\sum_{j=1}^p\kappa_jZ_j
$$
$$
A(\mathbf{Z},\alpha,\beta)=\frac{1}{\alpha}\sum_{i=1}^nX_i\log S_i=\frac{1}{\alpha}\sum_{i=1}^n X_i
    \log\!\left(\sum_{j=1}^p Z_j^{\alpha}\,\exp(-\beta c_{ij})\right)
$$

The authors look for vectors $\mathbf{Z}$ that maximise $f$ subject to the budget
constraint $\sum_j \kappa_j Z_j = \sum_i X_i$ and the non-negativity constraint
$Z_j \geq 0$ for all $j$. By the stochastic-stability result of Sandholm (2010,
§12.2), the global maximiser(s) of $f$ correspond to the most likely spatial
patterns when small random perturbations can occur.

### Gradient of the potential function

Differentiating $A$ with respect to $Z_j$ gives

$$
\frac{\partial A}{\partial Z_j}
  = \frac{1}{\alpha}\sum_{i=1}^n X_i \cdot \frac{\alpha Z_j^{\alpha-1}\,w_{ij}}{S_i}
  = \sum_{i=1}^n X_i\,\frac{Z_j^{\alpha-1}\,w_{ij}}{S_i}.
$$

Recognising that $Z_j^{\alpha-1} w_{ij} / S_i = P_{ij} / Z_j$, this becomes

$$
\frac{\partial A}{\partial Z_j} = \frac{1}{Z_j}\sum_{i=1}^n X_i P_{ij} = \frac{D_j}{Z_j}
$$
Adding the derivative of the linear term $-\kappa_j Z_j$, we obtain

$$
\frac{\partial f}{\partial Z_j}(\mathbf{Z},\alpha,\beta)
  = \frac{D_j - \kappa_j Z_j}{Z_j}
  = \frac{G_j(\mathbf{Z},\alpha,\beta)}{Z_j}
$$
The gradient $\nabla f$ vanishes at an interior point ($Z_j > 0$ for all $j$) if
and only if $G(\mathbf{Z}) = \mathbf{0}$, i.e. at a Harris--Wilson equilibrium.
Conversely, if $Z_j = 0$, the gradient is not directly defined, but the KKT
analysis below shows that boundary points can also be critical points of the
constrained problem.

Every interior maximum of $f$ is a Harris--Wilson equilibrium; the converse is
false in general. $G$ is not the gradient of $f$: $\partial f/\partial Z_j =
G_j/Z_j \neq G_j$. It follows that the Jacobian of $G$ is not the Hessian of
$f$, although the two are closely related at equilibrium.

### Hessian of the potential function

We introduce the softmax share $P_{ij} = Z_j^\alpha w_{ij}/S_i$ with
$S_i = \sum_k Z_k^\alpha w_{ik}$, and record its derivative with respect
to $Z_k$:
$$
\frac{\partial P_{ij}}{\partial Z_k}
  = \frac{\alpha}{Z_k}\,P_{ik}\bigl(\delta_{jk} - P_{ij}\bigr),
$$
where $\delta_{jk}$ is the Kronecker symbol.

#### The general Hessian

Recall that $\partial f/\partial Z_j = G_j/Z_j$, where
$G_j = D_j - \kappa_j Z_j$ is the disequilibrium gap.
Differentiating once more with respect to $Z_k$ yields the full Hessian
$H = \bigl(\partial^2 f/\partial Z_j\,\partial Z_k\bigr)$
at any interior point $\mathbf{Z} \gg 0$:

$$
H_{jk} =
\begin{cases}
\displaystyle
  -\frac{G_j}{Z_j^2}
  +(\alpha-1)\,Z_j^{\alpha-2}\sum_i X_i\,\frac{w_{ij}}{S_i}
  -\alpha\,Z_j^{2(\alpha-1)}\sum_i X_i\,\frac{w_{ij}^2}{S_i^2},
  & j = k,\\[10pt]
\displaystyle
  -\alpha\,Z_j^{\alpha-1}Z_k^{\alpha-1}
  \sum_i X_i\,\frac{w_{ij}\,w_{ik}}{S_i^2},
  & j \neq k.
\end{cases}
$$

#### Simplification at an interior equilibrium

At a Harris--Wilson equilibrium $G_j = 0$ for every active destination,
the term $-G_j/Z_j^2$ vanishes on the diagonal and the Hessian reduces to

$$
H_{jk}\big|_{G=0} =
\begin{cases}
\displaystyle
  (\alpha-1)\,Z_j^{\alpha-2}\sum_i X_i\,\frac{w_{ij}}{S_i}
  -\alpha\,Z_j^{2(\alpha-1)}\sum_i X_i\,\frac{w_{ij}^2}{S_i^2},
  & j = k,\\[10pt]
\displaystyle
  -\alpha\,Z_j^{\alpha-1}Z_k^{\alpha-1}
  \sum_i X_i\,\frac{w_{ij}\,w_{ik}}{S_i^2},
  & j \neq k.
\end{cases}
$$


The equilibrium is a strict local maximum of $f$ if and only if $H$ is
negative definite, i.e. $\lambda_{\max}(H) < 0$. The bifurcation threshold
corresponds to $\lambda_{\max}(H) = 0$.


### KKT Conditions and Equilibrium Structure

The Lagrangian of the constrained maximisation problem ($\max f$ subject to
$\sum_j \kappa_j Z_j = \sum_i X_i$, $Z_j \geq 0$) is

$$
\mathcal{L}(\mathbf{Z},\mu,\pmb{\lambda})
  = -f(\mathbf{Z},\alpha,\beta)
    - \sum_{j=1}^p \lambda_j Z_j
    + \mu\!\left(\sum_{i=1}^n X_i - \sum_{j=1}^p \kappa_j Z_j\right).
$$

The first-order KKT conditions give, for each $j$:

$$
\frac{\partial \mathcal{L}}{\partial Z_j}
  = -\frac{D_j}{Z_j} + \kappa_j - \lambda_j - \mu\kappa_j = 0,
  \qquad \lambda_j \geq 0, \quad \lambda_j Z_j = 0.
$$
Multiplying by $Z_j$ and summing over $j$, and using
$\sum_j D_j = \sum_i X_i = \sum_j \kappa_j Z_j$, one finds $\mu = -2$.
Therefore:

- If $Z_j > 0$: complementary slackness forces $\lambda_j = 0$, and the
  condition reduces to $D_j = \kappa_j Z_j$, i.e. $G_j = 0$.
- If $Z_j = 0$: $\lambda_j$ absorbs the residual and the condition is
  automatically satisfied.

Hence every KKT point of the constrained problem is a Harris--Wilson equilibrium
(possibly with inactive destinations), and conversely.


## References

Harris, B. (1964). "A model of locational equilibrium for retail trade", 
Penn-Jersey Transportation Study, Philadelphia, Pennsylvania (mimeo).

Harris, B., & Wilson, A. G. (1978). "Equilibrium Values and
Dynamics of Attractiveness Terms in Production-Constrained
Spatial-Interaction Models", Environment and Planning A: Economy and Space,
10(4), 371-388. <https://dx.doi.org/10.1068/a100371>

Jost, L. (2006), "Entropy and diversity", Oikos, 113: 363-375.
<https://dx.doi.org/10.1111/j.2006.0030-1299.14714.x>

Nystuen, J.D. and Dacey, M.F. (1961), "A graph theory
interpretation of nodal regions", Papers and Proceedings of the Regional
Science Association 7: 29–42. <https://dx.doi.org/10.1007/bf01969070>

Rihll, T., and Wilson, A. (1991), "Modelling settlement structures in
ancient Greece: new approaches to the polis", In City and Country in the
Ancient World, Vol. 3, Edited by J. Rich and A. Wallace-Hadrill, 58–95.
London: Routledge.

Wilson, A. (2008), "Boltzmann, Lotka and Volterra and spatial structural
evolution: an integrated methodology for some dynamical systems", J. R.
Soc. Interface.5865–871 <https://dx.doi.org/10.1098/rsif.2007.1288>

Osawa, M., Akamatsu, T., & Kogure, Y. (2025). Retail location as a
potential game: the Harris--Wilson model revisited. *Working paper*.

Sandholm, W. H. (2010). *Population Games and Evolutionary Dynamics*.
MIT Press.
