---
title: "Theoretical Background"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Theoretical Background}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(blvim)
```

## Static Models

### Spatial Interaction Models

Spatial interaction models aim to estimate flows between locations, for instance,
workers commuting from residential zones to employment zones. Models are built from:

-   A collection of $n$ origin locations described by some characteristics
$(X_i)_{1\leq i\leq n}$
-   A collection of $p$ destination locations described by some characteristics
$(Z_j)_{1\leq j\leq p}$
-   A collection of $n\times p$ characteristics of the difficulty of travelling
(in a broad sense) from origin $i$ to destination $j$, $c_{ij}$

The goal is to estimate flows $(Y_{ij})_{1\leq i\leq n, 1\leq j\leq p}$ from
origin locations to destination locations. A typical hypothesis is for the flows
to depend on characteristics as follows:

$$
Y_{ij}=f(X_i, Z_j, c_{ij}),
$$

for a well-chosen function $f$. The most well-known spatial interaction model is
the so-called *gravity model*, which takes the form:

$$
Y_{ij}\propto \frac{X_iZ_j}{c_{ij}^2},
$$

where $\propto$ means proportional to, $c_{ij}$ is supposed to be the distance
between origin $i$ and destination $j$, and the characteristics $X_i$ and $Z_j$
are assumed to be numerical.

### Constraints

Spatial interaction models can be instantiated with different constraints:

-   On production: the $(X_i)$ are interpreted as production values. Then the
model is *production-constrained* if $\sum_{j=1}^pY_{ij}=X_i$ for all $i$;
-   On capacity: the $(Z_j)$ are interpreted as destination capacities. Then the
model is *capacity-constrained* if $\sum_{i=1}^pY_{ij}=Z_j$ for all $j$;
-   On both: the *doubly-constrained* models are both production- and
capacity-constrained. One of the most common models used in practice is the
doubly-constrained gravity model.

### Maximum Entropy Models

In the late 1960s, Alan Wilson developed a collection of spatial interaction
models based on a maximum entropy principle. In those models, the flow is given by:

$$
Y_{ij} \propto X_iZ_j^{\alpha}\exp(-\beta c_{ij}),
$$

where $\alpha$ and $\beta$ are two parameters interpreted as follows:

-   $\alpha$ is a *return to scale* parameter: if $Z_j$ grows above $1$, its
actual attractiveness can increase in a super-linear way ($\alpha>1$) or in a
sub-linear way ($\alpha<1$);
-   $\beta$ acts as the *inverse of the scale of the costs* $c_{ij}$. If those
costs are distances, for instance, $\frac{1}{\beta}$ can be seen as a cut-off distance.

As with other spatial interaction models, maximum entropy models are constrained.
In `blvim`, we focus on production-constrained models and thus we have:

$$
Y_{ij} = \frac{X_iZ_j^{\alpha}\exp(-\beta c_{ij})}{\sum_{k=1}^pZ_k^{\alpha}\exp(-\beta c_{ik})},
$$

which ensures:

$$
\forall i,\ \sum_{j=1}^pY_{ij}=X_i.
$$

This model is implemented in `blvim` by the `static_blvim()` function.

## Dynamic Models

The models described above are *static*: they correspond to a frozen view of the
exchanges with no retroaction of the flow received at one destination location
on its attractiveness, for instance.

In the 1960s and 1970s, Harris and Wilson developed dynamic models that assume a form of
retroaction, initially in terms of an equilibrium state.

### Harris and Wilson Equilibrium Values

The main idea from Harris (1964) is that if we interpret incoming flows as
revenues, then the capacity/attractiveness of a destination location should be
adapted to those revenues (for production-constrained models). Essentially, this
could be written as:

$$
\forall j,\ \kappa_j Z_j= D_j,
$$

where the $\kappa_j$ are destination specific conversion factors between capacities
and costs, and where $D_j$ is the total flow incoming in $Z_j$, i.e.:

$$
\forall j,\  D_j=\sum_{i=1}^nY_{ij}.
$$

Applied to the maximum entropy production-constrained model, this leads to a
spatial interaction model defined implicitly as follows. We assume given the
production constraints $(X_i)_{1\leq i\leq n}$ and the cost matrix
$(c_{ij})_{1\leq i\leq n, 1\leq j\leq p}$, then the attractiveness $(Z_j)_{1\leq j\leq p}$
is a solution of the following collection of fixed-point equations:

$$
\forall j,\  \kappa_j Z_j=\sum_{i=1}^n\left(\frac{X_iZ_j^{\alpha}\exp(-\beta c_{ij})}{\sum_{k=1}^pZ_k^{\alpha}\exp(-\beta c_{ik})}\right),
$$

where $\alpha$, $\beta$, and $(\kappa_j)_{1\leq j\leq p}$ are fixed parameters.

From the $(Z_j)_{1\leq j\leq p}$, the flows are computed using the standard
production-constrained maximum entropy model.

### A Dynamic Model: The Boltzmann-Lotka-Volterra Model

The collection of fixed-point equations can be solved using different techniques,
for instance, Newton's method. The simplest solution is probably to perform fixed-point
iterations, as discussed in Harris & Wilson (1978). We start with some initial
values of the $Z$s, denoted $(Z^{0}_j)_{1\leq j\leq p}$, and then iterate:

$$
\forall j,\ Z_j^{t+1} := \frac{1}{\kappa_j}\sum_{i=1}^n\left(\frac{X_i{(Z^{t}_j)}^{\alpha}\exp(-\beta c_{ij})}{\sum_{k=1}^p{(Z^{t}_k)}^{\alpha}\exp(-\beta c_{ik})}\right),
$$

until convergence.

This iterative scheme can also be interpreted as a dynamic model in which the
attractiveness of a location is updated to match its incoming flow. The update
is not instantaneous, and thus an update parameter is introduced, $\epsilon$.
Then we start again with some initial values of the $Z$s and then iterate:

$$
\begin{align*}
\forall j,\ D_j^{t} &:= \sum_{i=1}^n\left(\frac{X_i{(Z^{t}_j)}^{\alpha}\exp(-\beta c_{ij})}{\sum_{k=1}^p{(Z^{t}_k)}^{\alpha}\exp(-\beta c_{ik})}\right),\\
\forall j,\ Z_j^{t+1} &:= Z_j^t +\epsilon\left(\frac{1}{\kappa_j} D^{t}_j-Z^{t}_j\right).
\end{align*}
$$

Using $\epsilon<1$, typically a small value of $0.01$, gives a full trajectory
for the $Z$s with the added value of preventing oscillation behaviour.

A variant of the dynamic model has been proposed by Wilson in 2008, replacing
the second update by a quadratic version:

$$
\forall j,\ Z_j^{t+1} := Z_j^t +\epsilon\left(\frac{1}{\kappa_j} D^{t}_j-Z^{t}_j\right)Z^{t}_j.
$$

Both solutions are implemented in `blvim()`.

### Diversity

A typical behaviour of the equilibrium solutions is a concentration of the
attractiveness on a subset of the destination locations. This can be measured by
the `diversity()` function. It uses the notion of diversity outlined in, for
example, Jost (2006). The main idea of this notion, rephrased in the context of
locations, is to find from the $Z$s an equivalent number of locations, say $m$,
for which the attractiveness would be evenly spread (that is, proportional to $\frac{1}{m}$).

If we normalise the $(Z_j)_{1\leq j\leq p}$ such that they sum to 1, we get the
probability distribution $\mathcal{P}=(p_j)_{1\leq j\leq p}$ over $\{1, \ldots, p\}$.
One way to interpret the notion of diversity is to measure how "random"
$\mathcal{P}$ is. Indeed, if it is concentrated on 1, for instance, with $p_1\simeq 1$,
then samples generated in an i.i.d. manner from $\mathcal{P}$ will not be diverse,
as they will mostly consist of the value 1. Then an *entropy* is a rather natural
diversity index. For instance, the Shannon entropy $H(\mathcal{P})=-\sum_{j}p_j\log p_j$
provides a good diversity index. Based on the principle outlined above, the
diversity of $\mathcal{P}$ is the number $m$ such that the diversity index of
$\mathcal{P}$ is equal to the diversity index of the uniform distribution on
$\{1, \ldots, m\}$. It can be shown that for the Shannon entropy, the diversity of
$\mathcal{P}$ is $\exp(H(\mathcal{P}))$. This holds also for all the Rényi entropies.

Notice that in `diversity()` the diversities are computed according to the incoming
flows, i.e., not using the $(Z_j)_{1\leq j\leq p}$ but rather the $D_j$ defined
above by
$$
\forall j,\  D_j=\sum_{i=1}^nY_{ij}.
$$


## Self-Exchange Configurations

In the general case, exchanges in spatial interaction models follow a bipartite
scheme: production locations send their productions to destination locations.
These locations are two distinct sets. The non-bipartite case is when production
and destination locations are identical. This does not change anything in the
definition of the models (static and dynamic), but new tools can be defined
specifically for this particular case.

### Location Domination and Terminals

These tools are based on notions proposed by J. D. Nystuen and M. F. Dacey (1961).
As any location is both a production and a destination location, one can compare
the outgoing flows to the incoming ones. In particular, Nystuen and Dacey focus
on the maximum output flow and on the total incoming flow (the $(D_j)_{1\leq j\leq p}$).
More precisely, they define the importance of location $j$ as $D_j$ and introduce
the notion of *domination*: a location $k$ is *dominated* if its importance is smaller
than the importance of the location to which it is sending its largest flow. In
equation:

$$
D_k < D_{\arg\max_{j} Y_{kj}}.
$$

A *terminal* is a location that is not dominated. Rihll and Wilson (1991) propose a
relaxed version in which a terminal is a location that receives a total incoming
flow larger than all its individual outgoing flows. Thus, location $k$ is dominated
according to this definition if:

$$
D_k < \max_{j}Y_{kj}.
$$

It is easy to see that a terminal in the Nystuen and Dacey definition is also a
terminal for Rihll and Wilson, but the reverse is false. The functions
`terminals()` and `is_terminal()` compute the terminals according to either
definition. In addition, the number of terminals can be seen as a diversity measure
and is thus included in the definitions supported by `diversity()`.

### Domination Graphs

In addition to extracting terminals from a non-bipartite model, Nystuen and Dacey
propose to analyse the so-called *nodal structure* of the flows. It consists of
the *nodal flows*, i.e. the links between each non-terminal location and the
location to which it is sending its largest flow. This defines a graph on the
locations which is very sparse (it has $p-t$ edges, where $p$ is the number of
locations and $t$ the number of terminals). The function `nd_graph()` computes
this graph according to both terminal definitions.

## Potential function from Osawa et al (2025)

Before going further we introduce simplified notation for the HW model :

$$
w_{ij}=w_{ij}(\beta)=\exp(-\beta c_{ij})
\qquad
S_i=S_i(\mathbf{Z},\alpha,\beta)=\sum_{k=1}^pZ_k^{\alpha}w_{ik}
\qquad
P_{ij}=\frac{Z_j^{\alpha}w_{ij}}{S_i}
$$
so that $P_{ij}$ is the share of origin $i$'s demand attracted to destination $j$.
The total flow received by destination $j$ is

$$
D_j(\mathbf{Z},\alpha,\beta) = \sum_{i=1}^n X_i P_{ij},
$$
and the Harris--Wilson fixed-point residual is

$$
G_j(\mathbf{Z},\alpha,\beta) = D_j(\mathbf{Z},\alpha,\beta) - \kappa_j Z_j,
$$
where $\kappa_j > 0$ is the capacity cost at destination $j$.
A **Harris--Wilson equilibrium** is a vector $\mathbf{Z}^* \geq 0$ satisfying
$G(\mathbf{Z}^*) = \mathbf{0}$.

### The potential function

In 2025 the team of researcher composed by Osawa, Akamatsu and Kogure introduced
a new concept: the HW model can be seen as a large population potential game.
This concept, which comes from game theory, allows us to see the HW model as a
single maximisation problem characterised by a potential function defined as:

$$
f(\mathbf{Z},\alpha,\beta)=A(\mathbf{Z},\alpha,\beta)-\sum_{j=1}^p\kappa_jZ_j
$$
$$
A(\mathbf{Z},\alpha,\beta)=\frac{1}{\alpha}\sum_{i=1}^nX_i\log S_i=\frac{1}{\alpha}\sum_{i=1}^n X_i
    \log\!\left(\sum_{j=1}^p Z_j^{\alpha}\,\exp(-\beta c_{ij})\right)
$$

The authors look for vectors $\mathbf{Z}$ that maximise $f$ subject to the budget
constraint $\sum_j \kappa_j Z_j = \sum_i X_i$ and the non-negativity constraint
$Z_j \geq 0$ for all $j$. By the stochastic-stability result of Sandholm (2010,
§12.2), the global maximiser(s) of $f$ correspond to the most likely spatial
patterns when small random perturbations can occur.

### Gradient of the potential function

Differentiating $A$ with respect to $Z_j$ gives

$$
\frac{\partial A}{\partial Z_j}
  = \frac{1}{\alpha}\sum_{i=1}^n X_i \cdot \frac{\alpha Z_j^{\alpha-1}\,w_{ij}}{S_i}
  = \sum_{i=1}^n X_i\,\frac{Z_j^{\alpha-1}\,w_{ij}}{S_i}.
$$

Recognising that $Z_j^{\alpha-1} w_{ij} / S_i = P_{ij} / Z_j$, this becomes

$$
\frac{\partial A}{\partial Z_j} = \frac{1}{Z_j}\sum_{i=1}^n X_i P_{ij} = \frac{D_j}{Z_j}
$$
Adding the derivative of the linear term $-\kappa_j Z_j$, we obtain

$$
\frac{\partial f}{\partial Z_j}(\mathbf{Z},\alpha,\beta)
  = \frac{D_j - \kappa_j Z_j}{Z_j}
  = \frac{G_j(\mathbf{Z},\alpha,\beta)}{Z_j}
$$
The gradient $\nabla f$ vanishes at an interior point ($Z_j > 0$ for all $j$) if
and only if $G(\mathbf{Z}) = \mathbf{0}$, i.e. at a Harris--Wilson equilibrium.
Conversely, if $Z_j = 0$, the gradient is not directly defined, but the KKT
analysis below shows that boundary points can also be critical points of the
constrained problem.

Every interior maximum of $f$ is a Harris--Wilson equilibrium; the converse is
false in general. $G$ is not the gradient of $f$: $\partial f/\partial Z_j =
G_j/Z_j \neq G_j$. It follows that the Jacobian of $G$ is not the Hessian of
$f$, although the two are closely related at equilibrium.

### Hessian of the potential function

We first record the derivative of the softmax share $P_{ij}$ with respect to $Z_k$:

$$
\frac{\partial P_{ij}}{\partial Z_k}
  = \frac{\alpha\,\delta_{jk}\,S_i\,Z_j^{\alpha-1}w_{ij}
         - \alpha\,Z_k^{\alpha-1}w_{ik}\,Z_j^{\alpha}w_{ij}}{S_i^2}
  = \frac{\alpha}{Z_k}\,P_{ik}\bigl(\delta_{jk} - P_{ij}\bigr),
$$

where $\delta_{jk}$ is the Kronecker symbol.

#### Off-diagonal entries ($j \neq k$)

Starting from $\partial f/\partial Z_j = D_j/Z_j - \kappa_j$ and differentiating
with respect to $Z_k \neq Z_j$:

$$
\frac{\partial^2 f}{\partial Z_j\,\partial Z_k}
  = \frac{1}{Z_j}\frac{\partial D_j}{\partial Z_k}
  = \frac{1}{Z_j}\sum_{i=1}^n X_i\,\frac{\partial P_{ij}}{\partial Z_k}
  = \frac{1}{Z_j}\sum_{i=1}^n X_i\,\frac{\alpha}{Z_k}\,P_{ik}(0 - P_{ij}),
$$
which gives

$$
H_{jk} = -\frac{\alpha}{Z_j Z_k}\sum_{i=1}^n X_i\,P_{ij}\,P_{ik}
        = -\alpha\,Z_j^{\alpha-1}Z_k^{\alpha-1}
          \sum_{i=1}^n X_i\,\frac{w_{ij}\,w_{ik}}{S_i^2},
\qquad j \neq k.
$$

#### Diagonal entries ($j = k$)

Differentiating $\partial f/\partial Z_j = D_j/Z_j - \kappa_j$ with respect to $Z_j$:

$$
\frac{\partial^2 f}{\partial Z_j^2}
  = -\frac{D_j}{Z_j^2} + \frac{1}{Z_j}\frac{\partial D_j}{\partial Z_j}.
$$
Using $\partial D_j/\partial Z_j = \sum_i X_i (\alpha/Z_j) P_{ij}(1-P_{ij})$:

$$
\frac{\partial^2 f}{\partial Z_j^2}
  = -\frac{D_j}{Z_j^2}
    + \frac{\alpha}{Z_j^2}\sum_{i=1}^n X_i\,P_{ij}(1-P_{ij}).
$$

At an **interior equilibrium** where $G_j = 0$, that is $D_j = \kappa_j Z_j$:

$$
H_{jj} = \frac{\alpha - 1}{Z_j^2}\sum_{i=1}^n X_i\,P_{ij}
         - \frac{\alpha}{Z_j^2}\sum_{i=1}^n X_i\,P_{ij}^2.
$$

#### Summary: the full Hessian

Collecting both cases, the Hessian matrix $H = (H_{jk})_{1 \leq j,k \leq p}$
evaluated at an interior equilibrium is

$$
H_{jk} =
\begin{cases}
\displaystyle
  (\alpha-1)\,Z_j^{\alpha-2}\sum_i X_i\,\dfrac{w_{ij}}{S_i}
  - \alpha\,Z_j^{2(\alpha-1)}\sum_i X_i\,\dfrac{w_{ij}^2}{S_i^2},
  & j = k, \\[12pt]
\displaystyle
  -\alpha\,Z_j^{\alpha-1}Z_k^{\alpha-1}
  \sum_i X_i\,\dfrac{w_{ij}\,w_{ik}}{S_i^2},
  & j \neq k.
\end{cases}
$$

### KKT Conditions and Equilibrium Structure

The Lagrangian of the constrained maximisation problem ($\max f$ subject to
$\sum_j \kappa_j Z_j = \sum_i X_i$, $Z_j \geq 0$) is

$$
\mathcal{L}(\mathbf{Z},\mu,\boldsymbol{\lambda})
  = -f(\mathbf{Z},\alpha,\beta)
    - \sum_{j=1}^p \lambda_j Z_j
    + \mu\!\left(\sum_{i=1}^n X_i - \sum_{j=1}^p \kappa_j Z_j\right).
$$

The first-order KKT conditions give, for each $j$:

$$
\frac{\partial \mathcal{L}}{\partial Z_j}
  = -\frac{D_j}{Z_j} + \kappa_j - \lambda_j - \mu\kappa_j = 0,
  \qquad \lambda_j \geq 0, \quad \lambda_j Z_j = 0.
$$
Multiplying by $Z_j$ and summing over $j$, and using
$\sum_j D_j = \sum_i X_i = \sum_j \kappa_j Z_j$, one finds $\mu = -2$.
Therefore:

- If $Z_j > 0$: complementary slackness forces $\lambda_j = 0$, and the
  condition reduces to $D_j = \kappa_j Z_j$, i.e. $G_j = 0$.
- If $Z_j = 0$: $\lambda_j$ absorbs the residual and the condition is
  automatically satisfied.

Hence every KKT point of the constrained problem is a Harris--Wilson equilibrium
(possibly with inactive destinations), and conversely.

## Stability Analysis and Bifurcation

### Local Stability via the Jacobian of $G$

The BLV dynamic model takes the form $\dot{Z}_j = \varepsilon\, G_j(\mathbf{Z},\alpha,\beta)$.
The local stability of an equilibrium $\mathbf{Z}^*$ is governed by the
**Jacobian of $G$**,

$$
J_G(\mathbf{Z}^*,\alpha,\beta)
= \left(\frac{\partial G_j}{\partial Z_k}\right)_{j,k}.
$$

The equilibrium $\mathbf{Z}^*$ is locally asymptotically stable if and only if
all eigenvalues of $J_G$ have strictly negative real part:

$$
\mathbf{Z}^* \text{ is locally stable}
\quad\Longleftrightarrow\quad
\lambda_{\max}\!\bigl(J_G(\mathbf{Z}^*,\alpha,\beta)\bigr) < 0.
$$

Using $\partial P_{ij}/\partial Z_k = (\alpha/Z_k)\,P_{ik}(\delta_{jk} - P_{ij})$,
the entries of $J_G$ are explicit. For $j \neq k$:

$$
\frac{\partial G_j}{\partial Z_k}
= -\frac{\alpha}{Z_k}\sum_{i=1}^n X_i\,P_{ij}\,P_{ik} \leq 0,
$$

and for the diagonal, using $D_j = \kappa_j Z_j^*$ at equilibrium:

$$
\frac{\partial G_j}{\partial Z_j}\bigg|_{\mathbf{Z}^*}
= \frac{\alpha}{Z_j^*}\!\left[D_j - \sum_{i=1}^n X_i P_{ij}^2\right] - \kappa_j
= (\alpha-2)\kappa_j - \frac{\alpha}{Z_j^*}\sum_{i=1}^n X_i P_{ij}^2.
$$

A **primary bifurcation** occurs at the critical value $\beta_c$ where
$\lambda_{\max}(J_G)$ first reaches zero along the dispersed branch.

### Behaviour of Collapsed Destinations

The iterative BLV scheme preserves positivity: if $Z_j^t > 0$ and
$\varepsilon < 1/\kappa_j$, then $Z_j^{t+1} = (1-\varepsilon\kappa_j)Z_j^t +
\varepsilon D_j^t > 0$. So $Z_j^* > 0$ at any finite iterate. However, for a
destination that has effectively collapsed, $Z_j^* \ll 1$ and
$P_{ij} \propto (Z_j^*)^\alpha \to 0$.

Examining the diagonal entry of $J_G$ for such a destination:

$$
\frac{\partial G_j}{\partial Z_j}\bigg|_{\mathbf{Z}^*}
= (\alpha-2)\kappa_j - \frac{\alpha}{Z_j^*}\sum_{i=1}^n X_i P_{ij}^2
\sim (\alpha-2)\kappa_j - \alpha\kappa_j (Z_j^*)^{\alpha-1}
\xrightarrow{Z_j^*\to 0}
(\alpha-2)\kappa_j,
$$

which remains **bounded**. The off-diagonal entries in the column and row of $j$,
of the form $-(\alpha/Z_k)\sum_i X_i P_{ij} P_{ik}$, vanish as $P_{ij} \to 0$.
Collapsed destinations therefore **decouple** cleanly in $J_G$: their rows and
columns both tend to zero, and $\lambda_{\max}(J_G)$ is determined entirely by
the remaining active destinations.

### Bifurcation Direction and the Dominant Eigenvector of $H$

When the primary bifurcation is approached, the loss of stability is not uniform
across destinations. The **direction** in which the system tips is encoded in the
eigenvector $\boldsymbol{\phi} \in \mathbb{R}^p$ associated with
$\lambda_{\max}(H)$, the dominant eigenvalue of the Hessian of $f$:

$$
H\,\boldsymbol{\phi} = \lambda_{\max}(H)\,\boldsymbol{\phi}.
$$

The sign of each component $\phi_j$ predicts the fate of destination $j$ at the
bifurcation:

- $\phi_j > 0$: destination $j$ **agglomerates** — its attractiveness grows
  above the current equilibrium as $\beta$ crosses the critical value.
- $\phi_j < 0$: destination $j$ **retracts** — its attractiveness falls below
  the current equilibrium, eventually collapsing toward zero.

The magnitude $|\phi_j|$ measures the **intensity** of the movement. This
eigenvector is computed in practice by `blv_split_direction()`, which restricts
the computation to the active subspace $\mathcal{A}^*$ (see below) to avoid
numerical issues near collapsed destinations.

**Remark on the diagonal approximation.** The diagonal $H_{jj}$ provides a
first-order approximation of $|\phi_j|$: destinations with the least-negative
$H_{jj}$ tend to have the largest $|\phi_j|$. However, the off-diagonal
structure of $H$ can reverse this ranking for destinations with strong spatial
interaction (e.g., a city may rank high by diagonal yet act as a retractor
because it shares its catchment with a dominant emerging city). The full
eigenvector is therefore always preferable to the diagonal alone.

#### Structural analogy with Rose (1998)

The BLV bifurcation structure bears a formal resemblance to the phase transitions
in the deterministic annealing framework of Rose (1998), developed independently
for vector quantisation and clustering. The formal correspondence at the level of
free energies is exact: setting $T=1$ in Rose's Lagrangian $F = D - TH$ and
using the effective distortion $d(i,j) = \beta t_{ji} - \alpha \log x_i$, one
recovers exactly the Harris--Wilson flows and the potential $f$ of Osawa et al.
(2025):

$$
f(\mathbf{x}) = -\frac{Q}{\alpha} F^*(\mathbf{x}) - \sum_i \kappa_i x_i,
$$

where $F^*$ is Rose's optimal Lagrangian value. Maximising $f$ (BLV) is
therefore equivalent to minimising $F^*$ (Rose) under the budget constraint.
Under this correspondence $\beta$ plays the structural role of the annealing
parameter ($\beta \to 0$: uniform dispersion; $\beta \to \infty$: full
agglomeration), while $T$ is held fixed at $1$.

**Where the analogy stops.** Rose's Theorem 1 — which states that a cluster
splits along the principal axis of its posterior covariance $C_{\mathbf{x}|y}$
when $T < 2\lambda_{\max}(C_{\mathbf{x}|y})$ — is proved specifically for the
**Euclidean distortion** $d = \|\mathbf{x}-\mathbf{y}\|^2$. The BLV distortion
$d(i,j) = \beta t_{ji} - \alpha \log x_i$ is non-Euclidean and **endogenous**
(it depends on the state $x_i$ through $\alpha \log x_i$). The phase-transition
condition must therefore be re-derived from the second-order optimality condition
of $F^*$ for this specific distortion, yielding:

$$
\sum_j \frac{Q_j}{Q}
\left[
  \alpha(\alpha-1)\sum_k p(k|j)\frac{\phi_k^2}{x_k^2}
  - \alpha^2 \left(\sum_k p(k|j)\frac{\phi_k}{x_k}\right)^{\!2}
\right] = 0,
$$

where $p(k|j) = x_k^\alpha e^{-\beta t_{jk}} / Z_j(x)$ is the Gibbs
distribution. This is the BLV analogue of Rose's condition
$\det(I - \frac{2}{T}C_{\mathbf{x}|y}) = 0$. Its spectral structure is
governed by the Hessian $H$ of $f$: the bifurcation direction $\boldsymbol{\phi}$
solving the above equation is precisely the dominant eigenvector of
$H_{\mathcal{A}^*}$. The posterior covariance $C_{\mathbf{x}|y}$ of Rose has
no direct counterpart in the BLV model — it is replaced by $H$ — and the two
objects are not formally equal.

## The Active Subspace

### Definition and Motivation

Even though $J_G$ itself remains well-behaved for collapsed destinations, the
**augmented system** used to locate bifurcation points exactly (see next section)
involves solving a linear system with $J_G$ as its principal block. For a
near-collapsed destination $j$, the diagonal entry
$(\alpha-2)\kappa_j - (\alpha/Z_j^*)\sum_i X_i P_{ij}^2$ contains a term
$\alpha/Z_j^* \to +\infty$ multiplied by $\sum_i X_i P_{ij}^2 \sim
(Z_j^*)^{2\alpha} \to 0$; their product $\sim (Z_j^*)^{2\alpha-1}$ may vanish,
but it does so through a near-cancellation that makes $J_G$ ill-conditioned near
collapsed destinations. To avoid this ill-conditioning, all subsequent analysis
— augmented systems, bifurcation direction, and type classification — is
restricted to the **active subspace** $\mathcal{A}^*$.

Two natural criteria define the active subspace, corresponding to the two
central objects of the BLV model. They lead to different algorithms with
different numerical properties.

**Via the Jacobian of $G$ (dynamic stability).** The active subspace is the
largest subset of destinations on which the restricted dynamics remain stable:

$$
\mathcal{A}^*_{J} = \underset{|\mathcal{A}|}{\arg\max}\,
\bigl\{
  \mathcal{A} \subseteq \{1, \ldots, p\}
  :
  \lambda_{\max}\!\bigl(J_{G_{\mathcal{A}}}\bigr) \leq 0
\bigr\}.
$$

In practice, $\mathcal{A}^*_J$ is identified by a **threshold on $Z_j^*$**: a
destination is considered active if and only if $Z_j^* > \tau$ for some small
threshold $\tau > 0$. This is justified by the decoupling property established
above — collapsed destinations ($Z_j^* \ll 1$) decouple from the active block
in $J_G$ — and is implemented in `blv_active()`.

**Via the Hessian of $f$ (potential concavity).** An interior maximum of $f$
requires $H$ to be negative semi-definite. The active subspace is the largest
subset on which the restricted Hessian remains concave:

$$
\mathcal{A}^*_{H} = \underset{|\mathcal{A}|}{\arg\max}\,
\bigl\{
  \mathcal{A} \subseteq \{1, \ldots, p\}
  :
  \lambda_{\max}(H_{\mathcal{A}}) \leq 0
\bigr\}.
$$

This criterion is implemented via a **greedy spectral algorithm** on $H$,
described in the next section (`blv_greedy()`).

**Relationship between the two criteria.** At a true interior equilibrium,
$H$ and $J_G$ are related by $H = \frac{1}{2}[\mathrm{diag}(1/Z)\,J_G +
J_G^\top\,\mathrm{diag}(1/Z)]$, so they share the same sign structure on the
active subspace and both criteria identify the same destinations as active. In
practice, they differ near collapsed destinations where $Z_j^* \approx 0$: the
threshold criterion on $Z_j^*$ is a direct proxy for $\mathcal{A}^*_J$, while
the greedy algorithm on $H$ constructs $\mathcal{A}^*_H$ by iterative spectral
elimination.

### Construction by Successive Elimination

#### Algorithm J: threshold on $Z_j^*$ (for $\mathcal{A}^*_J$)

The simplest approach to identifying $\mathcal{A}^*_J$ exploits the decoupling
property of $J_G$: because collapsed destinations have $P_{ij} \to 0$ and their
rows and columns in $J_G$ both vanish, $\lambda_{\max}(J_G)$ is driven entirely
by destinations with $Z_j^*$ bounded away from zero. The active subspace is
therefore approximated by:

$$
\mathcal{A}^*_J \approx \{j : Z_j^* > \tau\},
$$

for a small threshold $\tau > 0$ (default $\tau = 10^{-3}$ in `blv_active()`).
This approach is fast (no eigenvalue computation required) and numerically robust,
but it relies on a user-chosen threshold and does not provide a spectral
certificate on $J_{G_{\mathcal{A}^*}}$.

#### Algorithm H: greedy spectral elimination on $H$ (for $\mathcal{A}^*_H$)

The Hessian $H$ is symmetric and its eigenvalues are real, enabling a clean
spectral stopping criterion. Starting from the full index set
$\mathcal{A} = \{1, \ldots, p\}$:

1. Compute $\lambda_{\max}(H_{\mathcal{A}})$.
2. If $\lambda_{\max} \leq 0$, stop: $\mathcal{A}^* = \mathcal{A}$.
3. Otherwise, remove the index with the **largest diagonal entry** of $H$:
$$
j^* = \underset{j \in \mathcal{A}}{\arg\max}\; H_{jj}\big|_{\mathbf{Z}^*},
\qquad
\mathcal{A} \leftarrow \mathcal{A} \setminus \{j^*\},
$$
and return to step 1.

The algorithm terminates in at most $p - 1$ steps and is implemented in
`blv_greedy()`.

**Why the diagonal of $H$ as the elimination criterion?** At equilibrium:

$$
H_{jj} = \frac{\alpha - 1}{Z_j^{*2}}\sum_i X_i P_{ij}
         - \frac{\alpha}{Z_j^{*2}}\sum_i X_i P_{ij}^2.
$$

For a collapsed destination, $P_{ij} \sim (Z_j^*)^\alpha \to 0$, so $H_{jj}
\to 0^-$ — small and negative, making such destinations natural last candidates
for removal. For an emerging destination, the agglomeration term
$(\alpha-1)\sum_i X_i P_{ij}$ dominates the saturation term, giving $H_{jj} > 0$
— these are removed first, until the remaining submatrix is negative
semi-definite. Algorithm H is valid for all $\alpha > 1$.

**Why not use the diagonal of $J_G$?** For $1 < \alpha < 2$, the diagonal
entry of $J_G$ at a collapsed destination satisfies:

$$
\frac{\partial G_j}{\partial Z_j}\bigg|_{\mathbf{Z}^*}
\approx (\alpha - 2)\kappa_j < 0,
$$

so collapsed destinations have a *negative* diagonal entry in $J_G$ and would
never be selected for removal. Algorithm J via $J_G$ diagonal therefore fails
to construct $\mathcal{A}^*_J$ correctly for $\alpha < 2$, which covers the
empirically relevant range. The threshold approach avoids this issue entirely.

**Remark.** For $\alpha > 2$, the two diagonals agree in sign at collapsed
destinations (both positive), and both algorithms produce the same active
subspace. Algorithm H on the Hessian remains preferable in all cases because
$H$ is symmetric (real eigenvalues, numerically stable), while $J_G$ is not.

### The Reduced Stability Signal

On the active subspace $\mathcal{A}^*$, the restricted spectral radius
$\lambda_{\max}(H_{\mathcal{A}^*})$ remains bounded and interpretable
throughout the parameter range. Its evolution along a warm-start continuation
in $\beta$ reveals two types of events:

- **Primary bifurcation**: $\lambda_{\max}(H_{\mathcal{A}^*})$ reaches zero
  from below just before the first destination collapses, signalling the loss
  of stability of the fully dispersed equilibrium
  ($\mathcal{A}^* = \{1,\ldots,p\}$).
- **Secondary bifurcations**: each successive reduction of $|\mathcal{A}^*|$
  marks the collapse of one or more previously active destinations. Just before
  each event, $\lambda_{\max}(H_{\mathcal{A}^*})$ again approaches zero.

The same information is carried by $\lambda_{\max}(J_{G_{\mathcal{A}^*}})$ on
the active subspace, but $\lambda_{\max}(H_{\mathcal{A}^*})$ is preferred for
numerical stability and because $H$ is symmetric.

## Augmented Systems for Exact Bifurcation Location

All augmented systems below are formulated on the active subspace $\mathcal{A}^*$
of size $q = |\mathcal{A}^*|$, using the restricted Jacobian
$J_{G_{\mathcal{A}^*}} \in \mathbb{R}^{q \times q}$ and the restricted unknown
vector $\mathbf{Z}^*_{\mathcal{A}^*} \in \mathbb{R}^q$. This avoids the
ill-conditioning introduced by near-collapsed destinations.

### One Free Parameter: Locating $\beta_c$ Exactly

To locate the bifurcation point without relying on grid resolution, we apply the
standard augmented system of Kuznetsov (1995, Ch. 10). At a codimension-1
bifurcation, $J_{G_{\mathcal{A}^*}}(\mathbf{Z}^*,\beta_c)$ has a one-dimensional
kernel spanned by $\boldsymbol{\psi} \in \mathbb{R}^q$. The triple
$(\mathbf{Z}^*_{\mathcal{A}^*}, \beta_c, \boldsymbol{\psi})$ satisfies the
$(2q+1)$-dimensional square system:

$$
\begin{cases}
G_{\mathcal{A}^*}(\mathbf{Z}^*_{\mathcal{A}^*},\, \beta_c) = \mathbf{0},
  & (q \text{ equations})\\[6pt]
J_{G_{\mathcal{A}^*}}(\mathbf{Z}^*_{\mathcal{A}^*},\, \beta_c)\,
  \boldsymbol{\psi} = \mathbf{0},
  & (q \text{ equations})\\[6pt]
\boldsymbol{\psi}^\top \boldsymbol{\psi}_0 = 1,
  & (1 \text{ equation})
\end{cases}
$$

where $\boldsymbol{\psi}_0 \in \mathbb{R}^q$ is an approximate null vector from
a grid estimate, providing phase and sign normalisation. Kuznetsov (1995, Ch. 10)
proves that this Jacobian is non-singular at a generic codimension-1 point,
ensuring local quadratic convergence of Newton's method.

Note that $\boldsymbol{\psi}$ here is the **null vector of $J_G$** at the exact
bifurcation point, and encodes the direction of the emerging branch within
$\mathcal{A}^*$: destinations with $\psi_j > 0$ agglomerate; those with $\psi_j
< 0$ retract. This is distinct from the dominant eigenvector $\boldsymbol{\phi}$
of $H_{\mathcal{A}^*}$ introduced in the previous section, which approximates
this direction from the current equilibrium without requiring Newton's method.
Both vectors carry the same qualitative information on the bifurcation direction.

### Two Free Parameters: The Bifurcation Curve in $(\alpha,\beta)$-Space

Allowing both $\alpha$ and $\beta$ to vary, the augmented system becomes the
$(2q+1)$-dimensional system in the $(2q+2)$ unknowns
$(\mathbf{Z}^*_{\mathcal{A}^*}, \alpha, \beta, \boldsymbol{\psi})$:

$$
\begin{cases}
G_{\mathcal{A}^*}(\mathbf{Z}^*_{\mathcal{A}^*},\, \alpha,\, \beta) = \mathbf{0},\\[6pt]
J_{G_{\mathcal{A}^*}}(\mathbf{Z}^*_{\mathcal{A}^*},\, \alpha,\, \beta)\,
  \boldsymbol{\psi} = \mathbf{0},\\[6pt]
\boldsymbol{\psi}^\top \boldsymbol{\psi}_0 = 1.
\end{cases}
$$

This under-determined system (one free degree) defines a smooth curve in the
solution space. It is traced by **pseudo-arclength continuation** (Keller, 1977):
at each step a predictor advances along the tangent to the curve, and a corrector
Newton solve enforces the system augmented by one arclength constraint. The
projection onto the $(\alpha, \beta^{-1})$-plane gives the complete **phase
boundary** of the BLV model, extending the single-point analysis of Osawa et al.
(2025) to a full bifurcation portrait in two-parameter space.

## References

Harris, B. (1964). "A model of locational equilibrium for retail trade",
Penn-Jersey Transportation Study, Philadelphia, Pennsylvania (mimeo).

Harris, B., & Wilson, A. G. (1978). "Equilibrium Values and
Dynamics of Attractiveness Terms in Production-Constrained
Spatial-Interaction Models", Environment and Planning A: Economy and Space,
10(4), 371-388. <https://dx.doi.org/10.1068/a100371>

Jost, L. (2006), "Entropy and diversity", Oikos, 113: 363-375.
<https://dx.doi.org/10.1111/j.2006.0030-1299.14714.x>

Nystuen, J.D. and Dacey, M.F. (1961), "A graph theory
interpretation of nodal regions", Papers and Proceedings of the Regional
Science Association 7: 29-42. <https://dx.doi.org/10.1007/bf01969070>

Rihll, T., and Wilson, A. (1991), "Modelling settlement structures in
ancient Greece: new approaches to the polis", In City and Country in the
Ancient World, Vol. 3, Edited by J. Rich and A. Wallace-Hadrill, 58-95.
London: Routledge.

Wilson, A. (2008), "Boltzmann, Lotka and Volterra and spatial structural
evolution: an integrated methodology for some dynamical systems", J. R.
Soc. Interface.5865-871 <https://dx.doi.org/10.1098/rsif.2007.1288>

Kuznetsov, Y. A. (1995). *Elements of Applied Bifurcation Theory*. Springer.

Keller, H. B. (1977). "Numerical solution of bifurcation and nonlinear
eigenvalue problems", in *Applications of Bifurcation Theory*,
P. H. Rabinowitz (ed.), Academic Press, 359-384.

Osawa, M., Akamatsu, T., & Kogure, Y. (2025). Retail location as a
potential game: the Harris--Wilson model revisited. *Working paper*.

Rose, K. (1998). "Deterministic annealing for clustering, compression,
classification, regression, and related optimization problems",
Proceedings of the IEEE, 86(11), 2210-2239.
<https://dx.doi.org/10.1109/5.726788>

Sandholm, W. H. (2010). *Population Games and Evolutionary Dynamics*.
MIT Press.
