---
title: "Theoretical background"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Theoretical background}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(blvim)
```

## Spatial interaction models

Spatial interaction models try to estimate flows between locations, for instance 
workers commuting from residential zones to employment zones. Models are built from:

-   a collection of $n$ origin locations described by a some characteristics $(X_i)_{1\leq i\leq n}$
-   a collection of $p$ destination locations described by a some characteristics $(Z_j)_{1\leq j\leq p}$
-   a collection of $n\times p$ characteristics of the difficulty of travelling (in a broad sense) from origin $i$ to destination $j$, $c_{ij}$

The goal is to estimate flows $(Y_{ij})_{1\leq i\leq n, 1\leq j\leq p}$ from origin locations to destinations locations. A typical hypothesis is for the flows to depend on characteristics as follows

$$
Y_{ij}=f(X_i, Z_j, c_{ij}),
$$ 
for a well chosen function $f$. The most well known spatial interaction model
is the so called *gravity model* which takes the form
$$
Y_{ij}\propto \frac{X_iZ_j}{c_{ij}^2},
$$ 
where $\propto$ means proportional to, $c_{ij}$ is supposed to be the distance
between origin $i$ and destination $j$, and the characteristics $X_i$ and $Z_j$ 
are assumed to be numerical.

## Maximum entropy models
In the late 60s, Alan Wilson developed a collection of spatial interaction models
based of a maximum entropy principle. In those models, the flow is given by
$$
Y_{ij} \propto X_iZ_j^{\alpha}\exp(-\beta c_{ij}),
$$
where $\alpha$ and $\beta$ are two parameters interpreted as follows:

-   $Z_j$ is the attractiveness of the destination location $j$
-   $\alpha$ is then a *return to scale* parameter: if $Z_j$ grows above $1$, its
    actual attractiveness can increase in a super-linear way ($\alpha>1$) or in a
    sub-linear way $\alpha<1$
-   $\beta$ acts as the *inverse of the scale of the costs* $c_{ij}$. If those costs
    are distances, for instance, $\frac{1}{\beta}$ can be seen as a cut off distance. 
    
